### 2016-10-10 Projects/DelCo_2012/Report/DelCo_2012.tex
minor updates from JH review to CAFS

\begin{tabbing}
\hspace{1.5in} \= \\
CAF \> Compressed Air Foam \\
CAFS \> Compressed Air Foam System \\
Cal Poly \> California Polytechnic State University, San Luis Obispo \\
ESTC \> Delaware County Emergency Services Training Center \\
MCFRS \> Montgomery County (MD) Fire and Rescue Service \\
NIST \> National Institute of Standards and Technology \\
TFT \> Task Force Tips \\
\end{tabbing}

@@

This study, conducted by NIST, was part of a larger research project led by California Polytechnic State University (Cal Poly), San Luis Obispo in collaboration with the Montgomery County (MD) Fire and Rescue Service.  This study was supported in part by a DHS/FEMA Assistance to Firefighters Research and Development Grant.

@@

The Engineering Laboratory at the National Institute of Standards and Technology (NIST) has a research program aimed at improving the safety and effectiveness of firefighters through improved knowledge of fire behavior and fire fighting tactics. One of the objectives of this program is to examine non-traditional means of fire suppression. This series of experiments conducted by NIST for the California Polytechnic State University (Cal Poly), San Luis Obispo in collaboration with the Montgomery County (MD) Fire and Rescue Service complements the NIST research program on fire fighting tactics. These experiments are part of a larger research project led by Cal Poly with the purpose of investigating the capabilities and limitations of compressed air foam systems (CAFS) for structural fire fighting. The objective is to enhance the scientific knowledge base regarding the effectiveness and safety implications associated with the use of CAFS for structural fire fighting. This project was supported in part by a DHS/FEMA Assistance to Firefighters Research and Development Grant.

A number of fire departments throughout the US have acquired CAFS units for structural fire fighting, but many have limited or stopped the use of these CAFS for structural fire fighting. This is due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of compressed air foam (CAF) for structural fire fighting through the various tasks associated with this project, including:

@@

CAF generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved the water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters can see where they had applied the agent so that they could make a more effective application. In the wildland, water for fire fighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Finally, the compressed air adds energy to the water stream, providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that can be used to pre-treat homes and structures to protect them from an approaching fire or can be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires and also considered their use for structural fire fighting. Other departments across the country began to experiment with CAFS for use in structural fire fighting. In the few cases that were documented, the findings about the effectiveness of the CAFS were favorable. Yet, after decades of fire department trials, CAF has still not been widely incorporated into structural fire fighting tactics~\cite{USFA:CAFS,Lorh:2002}. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, had issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations with CAF inside a structure.

There has never been a comprehensive scientific study on the use of CAF for structural fire fighting. Fire departments have conducted pilot programs to examine implementation and effectiveness of CAF; foam suppliers and equipment manufacturers have conducted minimally instrumented experiments, typically in acquired structures; and test laboratories and universities have examined small sections of the CAF puzzle, but none of these efforts has resulted in the body of knowledge needed by the fire service~\cite{Madrzykowski:3,USFA:CAFS,Colletti:1,Boyd:1,Tafreshi:1,Madrzykowski:4,Colletti:2,Tafreshi:2,Lorh:2002,Colletti:3,Kim:2012}.

@@

\begin{table}[!ht]
\centering
\captionof{table}{Tests to Assess Fire Fighting Agents Compared to Water}\label{tab:agent_tests}
\begin{tabular}{lcc}
\toprule[1.5pt]
Test                                    & Similar Performance & Enhanced Performance \\
\midrule
Specific Heat                           &  \checkmark         &                      \\
Fuel Cooling and Penetration            &                     &  \checkmark          \\
Mass Retention                          &                     &  \checkmark          \\
Ignition Inhibition                     &                     &  \checkmark          \\
Drop Size                               &  \checkmark         &                      \\
Contact Angle                           &                     &  \checkmark          \\
Tire Fire Suppression                   &  \checkmark         &                      \\
Wood Crib Suppression                   &  \checkmark         &                      \\
Heptane Fire Suppression                &  \checkmark         &                      \\
Magnesium \& Titanium Fire Suppression  &  \checkmark         &                      \\
Smoke Generation                        &  \checkmark         &                      \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle; cooling and penetrating wood-based fuels; mass retention on the wood fuels; and ignition inhibition of the wood fuels.  When applied in the form of foam, the water agent solutions provide enhanced radiation exposure protection to a structure relative to plain water. The foams enhanced the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allowed the water to penetrate adsorbent fuels. These enhancements also make the agents valuable tools for mop-up or overhaul. The other experiments conducted during this study---specific heat, drop size, wood crib suppression, smoke generation, and Class D fire suppression---yielded similar results for both Class A foams and plain water.

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland fire fighting, CAFS is typically applied directly to the fuel from a remote location, typically a non-IDLH environment. In structural fire fighting, crews typically have to enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they have to mitigate the hazard of the hot gas layer or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior fire fighting with CAF are better understood.

>>>>

\begin{tabbing}
\hspace{1.5in} \= \\
BDP \> Bi-Directional Probe \\
CAF \> Compressed Air Foam \\
CAFS \> Compressed Air Foam System \\
Cal Poly \> California Polytechnic State University, San Luis Obispo \\
DHS \> Department of Homeland Security \\
FEMA \> Federal Emergency Management Administration \\
HF \> Heat Flux \\
IDLH \> Immediate Danger to Life and Health \\
ESTC \> Delaware County Emergency Services Training Center \\
MCFRS \> Montgomery County (MD) Fire and Rescue Service \\
NIST \> National Institute of Standards and Technology \\
OSB \> Oriented Strand Board \\
TC \> Thermocouple \\
TFT \> Task Force Tips \\
TJI \> Trust Joist I-Joist \\
UL FSRI \> UL Firefighter Safety Research Institute \\

\end{tabbing}

@@

This study, conducted by the National Institute of Standards and Technology (NIST), was part of a larger research project led by California Polytechnic State University (Cal Poly), San Luis Obispo in collaboration with the Montgomery County (MD) Fire and Rescue Service.  This study was supported in part by a DHS/FEMA Assistance to Firefighters Research and Development Grant.

@@

The Engineering Laboratory at the National Institute of Standards and Technology (NIST) has a research program aimed at improving the safety and effectiveness of firefighters through improved knowledge of fire behavior and fire fighting tactics. One of the objectives of this program is to examine non-traditional means of fire suppression. This series of experiments was conducted by NIST for the California Polytechnic State University (Cal Poly), San Luis Obispo in collaboration with the Montgomery County (MD) Fire and Rescue Service. It complements the NIST research program on fire fighting tactics. These experiments are part of a larger research project led by Cal Poly with the purpose of investigating the capabilities and limitations of compressed air foam systems (CAFS) for structural fire fighting. The objective is to enhance the scientific knowledge base regarding the effectiveness and safety implications associated with the use of CAFS for structural fire fighting. This project was supported in part by a Department of Homeland Security/Federal Emergency Management Administration (DHS/FEMA) Assistance to Firefighters Research and Development Grant.

A number of fire departments throughout the US have acquired CAFS units for structural fire fighting. Many have limited or stopped the use of these CAFS for structural fire fighting. This is due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of compressed air foam (CAF) for structural fire fighting through the various tasks associated with this project, including:

@@

CAF generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved the water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A aqueous solution into an aerated form (a blanket of bubbles) that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters can see where they had applied the agent so that they could make a more effective application. In the wildland, water for fire fighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Finally, the compressed air adds energy to the water stream, providing additional reach to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that can be used to pre-treat homes and structures to protect them from an approaching fire or can be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires and also considered their use for structural fire fighting. Other departments across the country began to experiment with CAFS for use in structural fire fighting. In the few cases that were documented, the findings about the effectiveness of the CAFS were favorable. Yet, after decades of fire department trials, CAF has still not been widely incorporated into structural fire fighting tactics~\cite{USFA:CAFS,Lorh:2002}. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, issued an order that the CAFS should not be used. The order was issued after incidents occurred where firefighters were burned while performing interior suppression operations with CAF inside a structure.

There has never been a comprehensive scientific study on the use of CAF for structural fire fighting. Fire departments have conducted pilot programs to examine implementation and effectiveness of CAF; foam suppliers and equipment manufacturers have conducted minimally instrumented experiments, typically in acquired structures; and test laboratories and universities have examined small sections of the CAF puzzle. None of these efforts has resulted in the body of knowledge needed by the fire service~\cite{Madrzykowski:3,USFA:CAFS,Colletti:1,Boyd:1,Tafreshi:1,Madrzykowski:4,Colletti:2,Tafreshi:2,Lorh:2002,Colletti:3,Kim:2012}.

@@

\begin{table}[!ht]
\centering
\captionof{table}{Tests to Assess CAF Fire Fighting Agent Compared to Water}\label{tab:agent_tests}
\begin{tabular}{lcc}
\toprule[1.5pt]
Test                                    & Similar Performance & Enhanced Performance \\
\midrule
Specific Heat                           &  \checkmark         &                      \\
Fuel Cooling and Penetration            &                     &  \checkmark          \\
Mass Retention                          &                     &  \checkmark          \\
Ignition Inhibition                     &                     &  \checkmark          \\
Drop Size                               &  \checkmark         &                      \\
Contact Angle                           &                     &  \checkmark          \\
Tire Fire Suppression                   &  \checkmark         &                      \\
Wood Crib Suppression                   &  \checkmark         &                      \\
Heptane Fire Suppression                &  \checkmark         &                      \\
Magnesium \& Titanium Fire Suppression  &  \checkmark         &                      \\
Smoke Generation                        &  \checkmark         &                      \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle; cooling and penetrating wood-based fuels; mass retention on the wood fuels; and ignition inhibition of the wood fuels.  When applied in the form of foam, the aqueous solutions provide enhanced radiation exposure protection to a structure relative to plain water. The foams enhanced the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allowed the water to penetrate adsorbent fuels. These enhancements also make the agents valuable tools for mop-up or overhaul. The other experiments conducted during this study---specific heat, drop size, wood crib suppression, smoke generation, and Class D fire suppression---yielded similar results for both Class A foams and plain water.

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland fire fighting, CAFS is typically applied directly to the fuel from a remote location, typically an environment that is not an immediate danger to life and health (non-IDLH environment). In structural fire fighting, crews typically have to enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they have to mitigate the hazard of the hot gas layer or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior fire fighting with CAF are better understood.
